---
title: "Slides Activity 3:"
subtitle: "Study planning"
toc: true
toc-expand: 2
format: html
---

# Using simulations to plan a study

Simulation isn't only useful for evaluating a method after data are in hand — it's also useful *before* a single participant is enrolled. If we can write down a plausible data generating mechanism (DGM) for our study, we can use it to ask design questions: If our hypothesized effect is real, how precisely will we estimate it, and how often will a single study actually show a clear difference, at a given sample size?

That question runs in two directions, and both come up in practice. If we're designing a prospective study, we usually get to ask "given the effect I hypothesize, how large a sample do I need?" But when we're reusing an existing data source — a claims database, a disease registry — the sample size is fixed by the data, not by us, and the useful question flips: "given the N I have, what's the smallest effect I could realistically detect?"

In this activity we revisit the prenatal vitamins / preterm birth / smoking example from Activities 1–2. Let's assume the effect is stronger (OR = 0.8) and use repeated simulation to see how precision (ASE, ESE) and our ability to detect a clear difference (power) change as sample size grows, and what the resulting confidence intervals actually look like. Then we'll turn the question around and ask what effect sizes are detectable at a couple of fixed, realistic sample sizes.

## Before you begin

Install the required packages.

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(kableExtra)
library(scales)
library(sandwich)
```

## Step 1. Outline the specific aims.

**Context:** We are planning a study of the effect of prenatal vitamin supplementation on the risk of preterm birth, worried about confounding by maternal smoking (unmeasured in our planned data source). Before we commit to a data source and sample size, we want to understand how sample size shapes what our study can tell us.

**Simulation Objectives:**

1. Assuming our hypothesized risk difference is correct, quantify how precision (ASE from both model-based and sandwich standard errors, and ESE) and the proportion of studies whose confidence interval would show a clear difference (power) change across a range of candidate sample sizes.

2. Turning that question around: at a couple of fixed, realistic sample sizes (as we might encounter with an existing registry or claims database), quantify the smallest effect size we could realistically detect.

## Step 2. Determine the appropriate data generating mechanism.

### Helper functions

First, create the "helper" functions we will use throughout our simulation.

```{r}
# The expit function is the inverse of the log-odds. It takes in a log-odds (x),
# applies the inverse-logit function and returns a probability
expit <- function(x) {
  return(exp(x) / (1 + exp(x)))
}

# The balancing intercept function, so we can re-derive the intercept whenever we
# change other parameters
derive_intercept <- function(py, prob_prenatal, prob_smoke,
                             or_smoke, or_prenatal){

  # marginal probability of treatment
  prob_prenatal <- prob_prenatal[1] * (1-prob_smoke) + prob_prenatal[2] * prob_smoke

  -log(1/py - 1) - or_prenatal*prob_prenatal - or_smoke*prob_smoke
}
```

### Function for DGM

This is the same DGM from Activities 1–2.

```{r}
dgm <- function(simid = 0,
                n = 500,
                prob_smoke = 0.2,
                prob_prenatal = c(0.85, 0.65),
                or_prenatal = log(0.8),
                or_smoke = log(3),
                prob_preterm = 0.1){

  # Dynamically derive the intercept term
  y_intercept <- derive_intercept(py = prob_preterm,
                                  prob_prenatal = prob_prenatal,
                                  prob_smoke = prob_smoke,
                                  or_smoke = or_smoke,
                                  or_prenatal = or_prenatal)

  dplyr::tibble(
    # Simulation number
    simid = simid,

    # Create n participants
    id = 1:n,

    # Randomly select smoking
    smoke = rbinom(n = n, size = 1, prob = prob_smoke),

    # Generate potential outcome under no treatment
    preterm_x0 = rbinom(n = n, size = 1,
                        prob = expit(y_intercept + (or_smoke * smoke))),

    # Generate potential outcome under treatment
    preterm_x1 = rbinom(n = n, size = 1,
                        prob = expit(y_intercept + (or_smoke * smoke) + or_prenatal)),

    # Generate observed treatment conditional on smoking
    p_prenatal = prob_prenatal[smoke + 1],
    p_prenatal_margin = prob_prenatal[1]*(1-prob_smoke) + prob_prenatal[2]*(prob_smoke),
    prenatal = rbinom(n = n, size = 1, prob = p_prenatal),

    # Assign actual outcome conditional on observed treatment
    preterm = preterm_x0*(1-prenatal) + preterm_x1*(prenatal),

    # Known IPT weight
    ipw = dplyr::if_else(
      prenatal == 1,
      p_prenatal_margin / p_prenatal,
      p_prenatal_margin / (1 - p_prenatal)
    )
  )
}
```

### Set the seed

ALWAYS before any random number generation!

```{r}
set.seed(918273)
```

### The hypothesized effect

Study planning happens *before* we have data, so instead of an empirical "truth" recovered from a huge dataset, we plug in our best guess (from prior literature, pilot data, or clinical judgment) for each DGM parameter. We can still generate one very large sample under those hypothesized parameters just to confirm what marginal risk difference they imply.

```{r}
truth <- dgm(n = 10000000)
true_rd <- mean(truth$preterm_x1) - mean(truth$preterm_x0)
true_rd
```

Notice this is a small effect — about a 1-percentage-point absolute reduction in preterm birth risk. Small, plausible drug effects like this are common in pharmacoepidemiology, and (as we'll see) they demand large studies.

## Step 3. Define the estimand of the analysis.

As in Activities 1–2, our estimand is the marginal risk difference, $E[Y^{x=1}] - E[Y^{x=0}]$: the risk of preterm birth had *everyone* in the target population taken prenatal vitamins vs. had *no one* taken them.

## Step 4. Identify the methods to be implemented in the study.

We will estimate the IPT-weighted risk difference (known weights), exactly as in Activity 2. Alongside the model-based standard error, we'll also compute the robust sandwich standard error, which accounts for the fact that we're using weights.

## Step 5. Specify the performance measures.

**Precision.** As in Activity 2, we'll track a few measures of how precisely we can pin down the risk difference:

- **ASE, model-based**: the average, across simulated samples, of the model-based standard error reported by the regression — what our software *tells us* about precision in a typical single sample.
- **ASE, sandwich**: the same idea, but averaging the robust sandwich standard error instead.
- **ESE** (empirical standard error): the standard deviation of the estimated risk differences across simulated samples — how much the estimate *actually* bounces around from sample to sample. If an ASE is close to the ESE, that standard error is trustworthy.

**Power.** Here we define power simply in terms of the confidence interval: the proportion of simulated samples whose 95% confidence interval (built from the model-based standard error) for the risk difference excludes zero (the value that would mean "no difference").

$$Power \approx P\Big(0 \notin \big[\widehat{RD} - 1.96\times\widehat{SE}(\widehat{RD}),\; \widehat{RD} + 1.96\times\widehat{SE}(\widehat{RD})\big]\Big)$$

In other words: if our hypothesized effect is real, in what proportion of studies of this size would the confidence interval actually show a clear difference, rather than being consistent with "no effect"?

## Simulating across a range of sample sizes

### Function to simulate one sample-size scenario

This function draws `nsim` replicate samples at a given `n`, fits the IPTW model to each, and returns the estimated risk difference, its model-based and sandwich standard errors, and whether the model-based 95% CI excludes zero — for every replicate.

```{r}
run_scenario <- function(n, nsim = 300) {
  purrr::map_dfr(1:nsim, function(i) {
    dat <- dgm(simid = i, n = n)
    
    unadj <- glm(preterm ~ prenatal, data = dat, family = binomial(link = "identity"))
    adj_known <- glm(preterm ~ prenatal, data = dat, 
                 weights = ipw, family = quasibinomial(link = "identity"))
    
    dplyr::tibble(
      estimator = c("Unadjusted", "Adjusted (Naive SE)", "Adjusted (Robust SE)"),
      
      # extract risk different from the models
      rd = c(coef(unadj)[2], coef(adj_known)[2], coef(adj_known)[2]),
      
      # extract the standard errors from the models
      se = c(sqrt(diag(vcov(unadj)))[2], 
             sqrt(diag(vcov(adj_known)))[2], 
             sqrt(diag(sandwich::vcovHC(adj_known)))[2])
    )
    }) |>
    dplyr::mutate(
      lcl = rd - 1.96*se,
      ucl = rd + 1.96*se,
      # Does this replicate's 95% CI exclude zero?
      excludes_zero = (lcl > 0) | (ucl < 0),
      # Does this replicate's CI include the truth?
      includes_truth = lcl <= true_rd & true_rd <= ucl
    )
}
```

### Run this across a grid of candidate sample sizes

```{r}
n_grid <- c(500, 1000, 2000, 5000, 10000, 20000, 40000)

results_by_n <- purrr::map_dfr(n_grid, function(n) {
  reps <- run_scenario(n, nsim = 300)
  reps |> 
    dplyr::summarize(
      n = n,
      power = mean(excludes_zero),
      ase = sqrt(mean(se^2)),
      ese = sd(rd),
      coverage = mean(includes_truth),
      .by = estimator
  )
})

results_by_n |>
  dplyr::arrange(estimator, n) |> 
  kableExtra::kable(col.names = c("Estimator", "N", "Power",  "ASE", "ESE", "95% confidence interval coverage"),
                     digits = c(0, 3, 4, 4, 4, 3))
```

### How does N affect precision (ASE, sandwich ASE, and ESE)?

```{r}
results_by_n |>
  tidyr::pivot_longer(cols = c(ase, ese), names_to = "measure", values_to = "value") |>
  dplyr::mutate(
    measure = dplyr::case_when(
      measure == "ese" ~ "ESE",
      measure == "ase" & estimator == "Adjusted (Naive SE)" ~ "ASE (model-based)",
      measure == "ase" & estimator == "Adjusted (Robust SE)" ~ "ASE (sandwich)")) |>
  distinct(n, value, measure) |>
  dplyr::filter(!is.na(measure)) |> 
  ggplot(aes(x = n, y = value, color = measure)) +
  geom_line() +
  geom_point() +
  scale_x_log10(breaks = n_grid, labels = scales::label_comma()) +
  theme_minimal() +
  labs(
    x = "Sample size (N, log scale)",
    y = "Standard error",
    color = "",
    title = "Precision of the risk difference estimate, by sample size",
    subtitle = "Comparing model-based ASE, sandwich ASE, and ESE as N grows"
  ) +
  theme(legend.position = "bottom")
```

### How does N affect power?

```{r}
results_by_n |> 
  dplyr::filter(estimator == "Unadjusted") |> 
  ggplot(aes(x = n, y = power)) +
  geom_line() +
  geom_point() +
  scale_x_log10(breaks = n_grid, labels = scales::label_comma()) +
  scale_y_continuous(limits = c(0, 1)) +
  theme_minimal() +
  labs(
    x = "Sample size (N, log scale)",
    y = "Proportion of samples whose 95% CI excludes zero",
    title = "How often would a study of this size show a clear difference?",
    subtitle = paste0("'True' RD = ", round(true_rd, 4))
  )
```

## Turning the question around: what effect could I detect with the data I have?

So far we've asked: "given a hypothesized effect, how big a sample do I need?" But a lot of pharmacoepidemiology works in the other direction — we're reusing an existing data source (an insurance claims database, a disease registry) whose size we don't get to choose. The useful planning question becomes: "given the N I'm stuck with, what's the smallest effect I could realistically detect?"

We can answer this with the exact same simulation machinery, just swapping which quantity is fixed and which one varies. Instead of looping over N at one hypothesized effect, we now fix N at a couple of realistic values and loop over a grid of hypothesized effect sizes.

### Function to simulate one effect-size scenario

```{r}
run_effect_scenario <- function(n, or_prenatal, nsim = 250) {
  purrr::map_dfr(1:nsim, function(i) {
    dat <- dgm(simid = i, n = n, or_prenatal = or_prenatal)

    fit <- glm(preterm ~ prenatal, data = dat,
               weights = ipw, family = quasibinomial(link = "identity"))

    dplyr::tibble(
      rd = coef(fit)[2],
      se = sqrt(diag(sandwich::vcovHC(fit)))[2]
    )
  }) |>
    dplyr::mutate(
      excludes_zero = (rd - 1.96*se > 0) | (rd + 1.96*se < 0)
    )
}
```

We use the sandwich standard error directly here, since the previous section showed it tracks the empirical standard error more closely than the model-based one.

### Run this across a grid of hypothesized effect sizes, at a couple of fixed sample sizes

```{r}
# Two data sources we don't get to redesign: a modest single-site registry,
# and a much larger insurance claims database.
fixed_n <- c(5000, 50000)
or_grid <- c(0.5, 0.6, 0.7, 0.8, 0.9, 0.95)

detectable_effect <- purrr::map_dfr(fixed_n, function(n) {
  purrr::map_dfr(or_grid, function(orp) {
    reps <- run_effect_scenario(n, or_prenatal = log(orp), nsim = 250)
    dplyr::tibble(n = n, or = orp, power = mean(reps$excludes_zero))
  })
})

detectable_effect |>
  kableExtra::kable(col.names = c("N", "Hypothesized OR", "Power"), digits = 3)
```

### Plot: power vs. hypothesized effect size, by sample size

```{r}
ggplot(detectable_effect, aes(x = or, y = power, color = factor(n))) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0.8, linetype = "dashed") +
  scale_y_continuous(limits = c(0, 1)) +
  scale_color_manual(values = c("#1b9e77", "#7570b3"), labels = scales::comma(fixed_n)) +
  theme_minimal() +
  labs(
    x = "Hypothesized odds ratio (smaller = stronger effect)",
    y = "Proportion of samples whose 95% CI excludes zero",
    color = "N",
    title = "What effect sizes could this study actually detect?",
    subtitle = "Weaker effects (closer to OR = 1) become undetectable sooner at smaller N"
  )
```

### The minimum detectable effect size, at each N

```{r}
#| warning: false
mdes <- detectable_effect |>
  dplyr::arrange(n, power) |>
  dplyr::summarize(
    min_detectable_or = approx(x = power, y = or, xout = 0.8)$y,
    .by = n
  )

mdes |>
  kableExtra::kable(col.names = c("N", "Smallest detectable OR (80% power)"), digits = 2)
```

At N = `r scales::comma(mdes$n[mdes$n == 5000])` — a plausible single-site registry — we could only reliably detect an odds ratio at or below about `r round(mdes$min_detectable_or[mdes$n == 5000], 2)`. At N = `r scales::comma(mdes$n[mdes$n == 50000])` — more like a claims database — that improves to about `r round(mdes$min_detectable_or[mdes$n == 50000], 2)`. This is exactly the question a study built on secondary data should ask *before* committing to it: given the N the data source hands you, is the effect you actually care about even detectable?

**Try it yourself:**

- What would be the sample size for 80% power if there was no relationship between smoking and treatment?
- What would be the impact to sample size if we had to incorporate another nuisance model: e.g., to correct for exposure misclassification? How could we incorporate in the simulation?
- How does the minimum detectable effect size change if the outcome (preterm birth) is rarer, e.g. `prob_preterm = 0.03` instead of 0.1? Does a rarer outcome make weaker effects harder or easier to detect at the same N?
